---
title: "p8105_hw2_pmd2137"
output: github_document
---

## Problem 0

```{r load libraries}
library(tidyverse)
library(readxl)

```

## Problem 1: (from solution posted)

This problem focuses on NYC Transit data; in particular, this CSV file contains information related to each entrance and exit for each subway station in NYC. If you’re not familiar with the NYC subway system, keeping a map in mind while looking at these data might help.

Read and clean the data; retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable (the ifelse or recode function may be useful).

```{r}
#read in the dataset, make sure all route cols are character data types
nyc_transit_data = read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
                            col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) %>%

  janitor::clean_names() %>%#run to clean dataset
  
  select(line, station_name, station_latitude, station_longitude, starts_with("route"), entry, exit_only, vending, entrance_type, ada) %>% #select specified colus
  
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE)) #convert binary data (yes/no) to logical
```

*Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?*

This data is not tidy. For the data to be tidy 'route' should be in long format.

```{r}
#distinct stations & lines 
nyc_transit_data %>% select(station_name, line) %>% distinct

#filter ADA compliance, distinct stations & lines 
nyc_transit_data %>% filter(ada == TRUE) %>% select(station_name, line) %>% distinct

#filter vending to NO
nyc_transit_data %>% filter(vending == "NO") %>% pull(entry) %>%  mean
```

*Answer the following questions using these data:*

1. How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St 8th Avenue; 125st Broadway; 125st Lenox); the distinct function may be useful here. *465 stations*

2. How many stations are ADA compliant? *84 stations*

3. What proportion of station entrances / exits without vending allow entrance? *37.72% of stations*


## Problem 2

This problem uses the Mr. Trash Wheel dataset, available as an Excel file on the course website.

Read and clean the Mr. Trash Wheel sheet: specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel use reasonable variable names omit rows that do not include dumpster-specific data round the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer)

```{r}
mr_trashwheel = read_excel("data/Trash_Wheel_Collection_Data.xlsx",sheet=1) %>%
  select (-...15,-...16) %>%
  janitor::clean_names() %>% #run to clean dataset
  mutate (sports_balls = as.integer(round(sports_balls,0)), source = "mr_trashwheel") #round sports_balls to integer, change data type

mr_trashwheel
```
Use a similar process to import, clean, and organize the data for Professor Trash Wheel. 
```{r}
#import professor trashwheel
prof_trashwheel = read_excel("data/Trash_Wheel_Collection_Data.xlsx",sheet=2) %>% #import data deom sheet 2
  janitor::clean_names() %>% #run to clean dataset, adjust column names
  mutate (sports_balls = as.integer(0), source = "prof_trashwheel") #add sports_balls variable as 0, add source variable
```

Combine this with the Mr. Trash Wheel dataset to produce a single tidy dataset. To keep track of which Trash Wheel is which, you may need to add an additional variable to both datasets before combining.

```{r}
#create joined dataset, and convert to tidy format
joined_trashwheel = rbind(mr_trashwheel,prof_trashwheel) %>%
  pivot_longer(c("plastic_bottles", "polystyrene","cigarette_butts", "glass_bottles", "grocery_bags", "chip_bags","sports_balls"), names_to = "trash_type", values_to = "amount_collected") #convert columns to tidy format (rather than do separately to each dataset then join)

```

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables. 

```{r}

# mr_trashwheel data consists of 548 observations measured across 15 variables including years, months, the amount of various types of garbage collected (cigarette butts, bags, balls, etc.). 

# prof_trashwheel data consists of 95 observations measured across 15 variables including years, months, the amount of various types of garbage collected (cigarette butts, bags, balls, etc.).

#joined_trashwheel is the tidy joined dataset. This dataset contains 643 unique observations with details recorded in tidy format. Column names for the tidy data are below.
names(joined_trashwheel)

```

*1. For available data, what was the total weight of trash collected by Professor Trash Wheel?* Total weight of trash was 380.24 tons

```{r}

#filter prof trashwheel,distinct dates & weights, sum all weights 
joined_trashwheel %>% filter(source == "prof_trashwheel") %>% select(dumpster,date,weight_tons) %>% distinct %>% pull(weight_tons) %>% sum

```

*2. What was the total number of sports balls collected by Mr. Trash Wheel in 2020?* 856 sports balls collected in 2020.

```{r}

#filter on mr trashwheel, year 2020, trash type as sports balls
joined_trashwheel %>% filter(source == "mr_trashwheel", year == 2020, trash_type == "sports_balls") %>% 
  select(dumpster,date,amount_collected) %>% distinct %>% #extract distinct counts for dates & dumpsters
  pull(amount_collected) %>% sum #extract sportsballs and sum

```

## Problem 3

This problem uses the FiveThirtyEight data; these data were gathered to create the interactive graphic on this page. In particular, we’ll use the data in pols-month.csv, unemployment.csv, and snp.csv. Our goal is to merge these into a single data frame using year and month as keys across datasets.

First, clean the data in pols-month.csv. 
- Use separate() to break up the variable mon into integer variables year, month, and day
- Replace month number with month name
- Create a president variable taking values gop and dem
- Remove prez_dem and prez_gop; and remove the day variable.

```{r}
pols_month = read.csv("data/fivethirtyeight_datasets/pols-month.csv") %>%
   separate(col=mon, into=c("year","month","day"), sep="-") %>% #separate date
   mutate(month = month.abb[as.numeric(month)], #convert month num to numeric, and then month
          year = substr(as.character(year),3,4), #reduce to 2 digit yr, for joining
          president = ifelse(prez_dem == 1, "dem", "gop")) %>% #code president as dem if dem, or gop if rep abbreviation
   select(-day, -prez_dem, -prez_gop) #drop day, prez_dem, prez_gop columns
```

Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

```{r}
snp = read.csv("data/fivethirtyeight_datasets/snp.csv")%>%
   mutate(date = as.Date(date, format = "%d/%m/%Y")) %>% 
   separate(col=date, into=c("year","month","day"), sep="-") %>% #separate date
   mutate(month = month.abb[as.numeric(month)], #convert month num to numeric, and then month abbreviation
          year = substr(as.character(year),3,4)) %>%
   select(-day)
```

Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

```{r}
unemployment = read.csv("data/fivethirtyeight_datasets/unemployment.csv") %>%
  mutate(year = substr(as.character(Year),3,4)) %>%
  pivot_longer(c("Jan","Feb", "Mar", "Apr", "May","Jun","Jul","Aug","Sep","Oct","Nov","Dec"), names_to = "month", values_to = "unemploy") %>%
  select(-Year)
```

Join the datasets by merging snp into pols, and merging unemployment into the result.

```{r}
merged_538_data = merge(pols_month,snp,by=c("year","month")) %>% merge(unemployment,by=c("year","month"))
```

*Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).*


The *pols_month* dataframe contains data about the number of various types of politicians from each major political party. The dataset has 822 entries, across 9 variables including. After wrangling, the dataset contains variables for: year (range 1947-2015), month, presidential political party, the number of republican and democratic governors, the number of republican and democratic senators, the number of republican and democratic representatives. 

The *snp* dataframe contains data about the closing S&P stock market price across several months and years. The dataset has 822 entries, across 2 variables - date, and closing. After wrangling, the dataset contains variables for: year (range 1950-2015), month, and closing value. 

The *unemployment* dataframe contains data about the percentage of unemployment across months and years. The dataset has 68 rows representing years, across 12 columns representing months. After wrangling, the dataset contains variables for: year (range 1948-2015), month, unemployment_%. 

