---
title: "p8105_hw2_pmd2137"
output: github_document
---

## Problem 0

```{r load libraries}
library(tidyverse)
library(readxl)
```

## Problem 1:

This problem focuses on NYC Transit data; in particular, this CSV file contains information related to each entrance and exit for each subway station in NYC. If you’re not familiar with the NYC subway system, keeping a map in mind while looking at these data might help.

Read and clean the data; retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable (the ifelse or recode function may be useful).

```{r}
#read in the dataset, make sure all route cols are character data types
nyc_transit_data = read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
                            col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) %>%

  janitor::clean_names() %>%#run to clean dataset
  
  select(line, station_name, station_latitude, station_longitude, starts_with("route"), entry, exit_only, vending, entrance_type, ada) %>% #select specified colus
  
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE)) #convert binary data (yes/no) to logical
  
nyc_transit_data

```

## Problem 2

This problem uses the Mr. Trash Wheel dataset, available as an Excel file on the course website.

Read and clean the Mr. Trash Wheel sheet: specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel use reasonable variable names omit rows that do not include dumpster-specific data round the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer)

```{r}
getwd()

#trashwheel_dataset = read_csv("data/Trash Wheel Collection Data.csv")

```

Use a similar process to import, clean, and organize the data for Professor Trash Wheel, and combine this with the Mr. Trash Wheel dataset to produce a single tidy dataset. To keep track of which Trash Wheel is which, you may need to add an additional variable to both datasets before combining.

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables. For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of sports balls collected by Mr. Trash Wheel in 2020?

```{r}

```

## Problem 3

This problem uses the FiveThirtyEight data; these data were gathered to create the interactive graphic on this page. In particular, we’ll use the data in pols-month.csv, unemployment.csv, and snp.csv. Our goal is to merge these into a single data frame using year and month as keys across datasets.

First, clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.

Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

Join the datasets by merging snp into pols, and merging unemployment into the result.

Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).

